The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-02-18 Paper.Rmd
minor change
edited methods and background and some of the fig for extinction risk stuff

We addressed Q3 using both models and data. An initial 
idea was presented in the Introduction that
the distribution (through time) of a spatially averaged quantity should be influenced by the 
nature of the dependence between the local quantities being averaged, including their
copula structure and tail dependence. For instance, suppose $E_i(t)$ are ecological
variables measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and suppose 
the distribution and temporal dynamics of the spatial average $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
strongly right-tail dependent for all or most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in all or most locations.
This should tend to increase the skewness of the distribution through time of the
spatial average. Similarly, left-tail dependence between local variables
will tend to decreases skewness of the spatial mean. Skewness of the 
spatial mean corresponds to the "spikiness" of the spatial-mean time series or
its tendency to crash to low levels. 
The spatial-mean time series and its skewness may be quantities of 
principal importance for 
cases such as pest or resource abundance or CO$_2$ or methane flux.

We tested these ideas using the focal multivariate datasets of our earlier analyses:
green spruce aphid abundance data, leaf-curling plum aphid first flight data, \emph{Ceratium
furca} abundance data, and methane flux data (see Data and Table \ref{tab_multivar_summary}).
For each dataset we calculated the spatial mean, and then the skewness (through time) of that
mean. Then, for each dataset separately, we compared the value obtained 
to a distribution of values 
of the same quantity calculated for each of 10000 surrogate datasets that were 
produced from the empirical data by randomizing
it in a special way to have the copula structure of a multivariate-normal distribution
(Appendix \ref{SM-Surrog_multi}). The surrogate/randomized datasets had exactly the same
marginal distributions (i.e., distributions of values for each sampling location)
as the original data. 
They also had very similar Spearman correlations between pairs of sampling locations
as the original data. But the surrogates had multivariate normal copula structure. 
Our comparisons therefore tested the null hypothesis that the skewness of the spatial 
mean took values on the empirical data no different from what one would expect if the copula
structure of the data were multivariate normal, but the data were otherwise statistically
unchanged. Significant differences indicate that copula structure contributed 
to the skewness of the spatial mean, i.e., to its "spikiness."

@@

<!--***DAN: 
1) Shya put concepts on why we would expect lt dependence of noise to mean
more ext risk in a metapop context
2) then methods for how we test that (present the model, say what sims you did-->

On a second thought, we addressed the question how asymmetry in 
tail-dependence of environmental
variable could influence the extinction risk from a meta-population
context. We hypothesized to get higher 
extinction risk if the environmental noises from multiple patches
were more correlated to their smaller values 
(resulting left-tail dependence
in copula structure) than larger. This idea came from the context that
"bad" years (e.g. extreme cold) would have adverse affect 
on all metapopulation patches simultaneously, so there would 
not be enough neighboring-populations to rescue the "extinct"
patches. But in "good" years as the environmental variables 
from multiple locations were not correlated strongly, some of
extinct metapopulation-patches could be rescued by dispersal from
other patches. We tested our hypothesis in the following section.

@@

<!--***DAN: Shya to insert extinction risk results with one fig-->
We tested our hypothesis of asymmetric copula-effect on extinction-risk
as discussed in
earlier section, with Lewinton-Cohen metapopulation
model (Eq. \ref{eq.extrisk_LC}), where population density 
$P$ could update exponentially depending on its intrinsic rate $r$.
$\epsilon(t)$ represent environmental noise; they are standard-normal 
random variables that are independent for distinct times, t, but 
that were made to exhibit either extreme left tail or right tail dependence
across patches.
We considered metapopulation patches were arranged in
a linear chain fashion
and the population may disperse to its nearerst
two neighbors (local dispersion) or to any patches (global dispersion); for code see [`mydispersal.R`](https://github.com/sghosh89/BIVAN/blob/master/mydispersal.R).

\begin{equation}\label{eq.extrisk_LC}
P(t+1) = \exp\left [  r + \epsilon(t) \right ] P(t)
\end{equation}

We simulated Eq. \ref{eq.extrisk_LC} with local dispersal for $r=0$ and calculated extinction risk after 25 years
from $10000$ simulations for both left-tail and right-tail dependent environmental noise across $5$ patches. 
Whenever population, at any patch after each year, went below the extinction threshold ($=1$) we considered 
that patch for that time went extinct. We found, as we
hypothesized, left tail dependent environmental fluctuations increased extinction risk of metapopulations than the right-tail
dependent ones (Fig. \ref{fig_extrisk_lcohen_local}). When number of patches increased from $5$ to $25$, extinction risk 
decreased in both cases. We repeated our simulations using same parameters with global dispersal among patches 
and got similar results (see Fig. \ref{SM-fig_extrisk_lcohen_global}).

\begin{figure}[!h]
\textbf{ \hspace{6.6 cm} (A) \hspace{4.2 cm} (B)} \\
\vspace{-0.8 cm}
\begin{center}
\includegraphics[width=10 cm]{./Results/ext_risk_copula/lcohen_ext_risk_local_disp.pdf}
\caption{Extinction risk for Lewinton-Cohen metapopulation model was calculated after $25$ years with local dispersal
among two nearest neighbors (Eq. \ref{eq.extrisk_LC}, $r=0$). $(A)$ Left-tail dependent environmental noise showed higher 
extinction risk than $(B)$ right-tail dependent one, where both case showed extinction risk decreased with 
increasing number of patches from $numlocs =5$ to $25$.\label{fig_extrisk_lcohen_local}}
\end{center}
\end{figure}

>>>>

We addressed Q3 using both models and data. An initial 
idea was presented in the Introduction that
the distribution (through time) of a spatially averaged quantity should be influenced by the 
nature of the dependence between the local quantities being averaged, including their
copula structure and tail dependence. For instance, suppose $E_i(t)$ are ecological
variables measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and suppose 
the distribution and temporal dynamics of the spatial average $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
strongly right-tail dependent for all or most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in all or most locations.
This should tend to increase the skewness of the distribution through time of the
spatial average. Similarly, left-tail dependence between local variables
will tend to decreases skewness of the spatial mean. Positive skewness of the 
spatial mean corresponds to the "spikiness" of the spatial-mean time series; negative 
skewness, if it occurs,  corresponds to a 
tendency to crash to low levels but quickly recover. 
The spatial-mean time series and its skewness may be quantities of 
principal importance for 
cases such as pest or resource abundance or CO$_2$ or methane flux.

We tested these ideas using the focal multivariate datasets of our earlier analyses:
green spruce aphid abundance data, leaf-curling plum aphid first flight data, \emph{Ceratium
furca} abundance data, and methane flux data (see Data and Table \ref{tab_multivar_summary}).
For each dataset we calculated the spatial mean, and then the skewness (through time) of that
mean. Then, for each dataset separately, we compared the value obtained 
to a distribution of values 
of the same quantity calculated for each of 10000 surrogate datasets. Surrogate 
datasets were produced from the empirical data by randomizing
it in a special way so that surrogate/randomized dataset then had the 
copula structure of a multivariate-normal distribution
(Appendix \ref{SM-Surrog_multi}). The surrogate datasets had exactly the same
distributions of values for each sampling location as the original data. 
They also had very similar Spearman correlations between pairs of sampling locations
as the original data.  
Our comparisons therefore tested the null hypothesis that the skewness of the spatial 
mean took values on the empirical data no different from what one would expect if the copula
structure of the data were multivariate normal, but the data were otherwise statistically
unchanged. Significant differences indicate that non-N copula structure in the data 
contributed to the skewness of the spatial mean, i.e., to the "spikiness" of that
time series.

@@

We also examined whether the nature of the
tail dependence of an environmental variable across space could influence the 
extinction risk of a metapopulation. We hypothesized that environmental noises
exhibiting greater left-tail dependence across different habitat patches would 
cause higher metapopulation extinction risks because then bad years for the 
component populations would occur simultaneously in many patches,
reducing the potential for rescue effects. 
We provided a test of the conceptual validity of this hypothesis using a metapopulation
extension of the Lewinton-Cohen model,
$\vec{P}(t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of populations, the $i$th component $P_i(t)$ representing
the population in the $i$th habitat patch ($i=1,\ldots,N$) at time $t$. 
The $N \times N$ matrix $\lambda(t)$ was diagonal with $i$th diagonal entry 
$\exp(r+\epsilon_i(t))$. Here $r$ is an intrinsic growth rate; we used $r=0$. The
$\epsilon_i(t)$ represent environmental noises in each patch. They were standard-normally 
distributed in each patch, were independent through time, and showed between-patch
Spearman correlations X and Pearson correlations Y in all simulations, 
<!--***Shya: Please fill in X and Y. If it was the Kendall that was used, please 
change Spearman to Kendall. If anything is unclear here in what you should do, please ask. 
If you don't know how to answer these questions, let me know and remind me where the 
code is that you used to generate the noise. Actually, even if you can do this yourself,
please insert a comment below this one indcating where the code is for model and noise-->
but were made to exhibit right- or left-tail dependence between patches in
different simulations (Appendix \ref{SM-ext_risk_model}). The $N \times N$
matrix $D$ was a dispersal matrix. For simplicity, we considered habitat patches 
to be arranged in a line, with populations dispersing either to nearest neighbors only,
or globally, in different simulations (Appendix \ref{SM-ext_risk_model}). The dispersal 
rate was $d$. After each time step, if the population in a patch was below $1$, it was
set to $0$. We simulated the model 10000 times for each combination of parameters,
starting from initial populations in each patch of X, 
<!--***DAN: Shya, please fill in X-->
and calculated extinction risk after $25$ time steps.

@@

<!--***DAN: Dan has not edited the results for the extinction risk stuff yet-->
 We found, as we
hypothesized, left tail dependent environmental fluctuations increased extinction risk of metapopulations than the right-tail
dependent ones (Fig. \ref{fig_extrisk_lcohen_local}). When number of patches increased from $5$ to $25$, extinction risk 
decreased in both cases. We repeated our simulations using same parameters with global dispersal among patches 
and got similar results (see Fig. \ref{SM-fig_extrisk_lcohen_global}).

\begin{figure}[!h]
\textbf{ \hspace{6.6 cm} (A) \hspace{4.2 cm} (B)} \\
\vspace{-0.8 cm}
\begin{center}
\includegraphics[width=10 cm]{./Results/ext_risk_copula/lcohen_ext_risk_local_disp.pdf}
\caption{Extinction risk for Lewinton-Cohen metapopulation model was calculated after $25$ years with local dispersal
among two nearest neighbors (Eq. \ref{eq.extrisk_LC}, $r=0$, $d=$ degree of dispersal). 
$(A)$ Left-tail dependent environmental noise showed higher 
extinction risk than $(B)$ right-tail dependent one, where both case showed extinction risk decreased with 
increasing number of patches from $numlocs =5$ to $25$.\label{fig_extrisk_lcohen_local}}
\end{center}
\end{figure}
<!--***DAN: Shya, can you please change this figure as follows:
1) Make it only go to d=0.5. I think reviewers will complain that d close to 1 is a weird and 
unrealistic. Anyway, results are the same for larger d, so we are not hiding anything
2) the main comparison is between right and left tail dependence, so make the left panel show 
the numlocs=5 case, with results for both right- and left-tail dependence of the noise shown. 
And make the right panel show the numlocs=25 case with results for both right- and left-tail
dependence of the noise shown.
3) Right now you use grey dots on one panel and black on the other, but you can just use black 
on both.
4)  Use N instead of numlocs - in papers we use variable names like in math, not like in code.
-->
